---
title: "Econometrics Project"
author: "Steven Trochez, Habiba Sultana"
date: "12/21/2019"
output: html_document
---
---
title: "Econometrics Project"
author: "Steven Trochez, Habiba Sultana"
date: "12/20/2019"
output: html_document
---
#Health condition and behaviors Analysis
This analysis aims at finding a relationship that links personal metrics (i.e. Race, Sex, Region) with health disparities across the United States. The use of multilinear regression is used to find a linear relationship between the predictors and targets (Chronic Disease, Smoking Habits, Drinking Habits, BMI levels).

The data is extracted from IPUMS/NHIS database. Courtsy of Lynn A. Blewett, Julia A. Rivera Drew, Miriam L. King and Kari C.W. Williams. IPUMS Health Surveys: National Health Interview Survey, Version 6.4 [dataset]. Minneapolis, MN: IPUMS, 2019. https://doi.org/10.18128/D070.V6.4 http://www.nhis.ipums.org


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r cars}
#Installing the neccesarry packages:
library(data.table)
library(dplyr)
library(igraph)
library(DataExplorer)
library(jtools)
library("naniar")
library(ggplot2)
library("ipumsr")
library(AER)
```
#Exporting and importing Data:
The data is accessible through the website (http://www.nhis.ipums.org). The Variables are chosen from a variaty of different options. The variables were chosen based on the requirments of the general question of the analysis.
```{r}
# Import the data file from local directory:
ddi <- read_ipums_ddi("nhis_00007.xml")
data <- read_ipums_micro(ddi)

#Summary of Data:
glimpse(data)
summary(data) 
```
#Data Cleaning:
There are three types of missing data (NIU, unknown-refused, uknown-not ascertined, unknown-don’t know), the data will be subsetted discarding these values.
```{r}
## Cleaning Data (Removing NA), Creating  tidy data:
data1<-replace_with_na(data, replace = list(EDUC=c(000, 996, 997, 998,999),
                                              RACEA=c(900,970,980,990),
                                              REGION=c(08,09),
                                              ASTHMAEV=c(0,7,8,9),
                                              CHEARTDIEV=c(0,7,8,9),
                                              ALCSTAT1=c(0,9),
                                              CIGSDAY=c(00,96,97,98,99),
                                              VISITYRNO=c(00,96,97,98,99),
                                              HINOTCOV=c(00,09),
                                              HEARTCONEV=c(0,7,8,9),
                                              STROKEV=c(0,7,8,9),
                                              ARTHLIMIT=c(0,7,8,9),
                                              DIABETICAGE=c(00,96,97,98,99),
                                              HEALTH=c(0,7,8,9),
                                              ALCSTAT1=c(0,9),
                                              SMOKFREQNOW=c(0,7,8,9),
                                              ASAD=c(0,6,7,8,9),
                                              MOD10DTP=c(0,7,8,9),
                                              HRSLEEP=c(00,97,98,99)))
                                            
```
#SUMMARAZING THE DATA: 
```{r} 
summary(data1)
attach(data1)
```
#CREATING MORE DESCRIPTIVE CATEGORIES ACCORDING TO THE IPUMS/NHIS CODEBOOK:
```{r}
#REGION:
REGION <- as.factor(REGION)
levels(REGION) <- c("Northeast","Midwest","South","West")
#AGE:
age<-as.numeric(AGE)
#SEX:
female <- as.numeric(SEX == 2)
#EDUCATION:
educ_nohs <- as.numeric(EDUC == 200)
educ_hs <- as.numeric((EDUC > 204) | (EDUC == 301))
educ_smcoll <- as.numeric(EDUC > 400) & (EDUC == 401)
educ_as <- as.numeric((EDUC > 402) & (EDUC < 403))
educ_bach <- as.numeric(EDUC == 500)
educ_adv <- as.numeric((EDUC > 600) & (EDUC < 603))
#RACE:
white <- as.numeric(RACEA == 100)
black <- as.numeric(RACEA == 200)
native <- as.numeric(RACEA == 300)
asian <- as.numeric(RACEA == 410)
#HEALTH:
HEALTH <- as.numeric(HEALTH == 1) | (HEALTH == 2) | (HEALTH == 3) | (HEALTH == 4) | (HEALTH == 5)
#BMI:
BMI <-as.numeric(BMI)
#DOCTOR VISITS:
VISITYRNO <-as.numeric(VISITYRNO > 10) & (VISITYRNO < 40)
#HEALTH INSURANCE:
HINOTCOV <-as.numeric(HINOTCOV == 1) | (HINOTCOV == 2)
#ASTHMA:
ASTHMAEV <-as.numeric(ASTHMAEV == 2)
#HEART DISEASE:
CHEARTDIEV <-as.numeric(CHEARTDIEV == 2)
#HIGH COLESTEROL:
CHOLHIGHEV <-as.numeric(CHOLHIGHEV == 2)
#HEART CONDITION:
HEARTCONEV <-as.numeric(HEARTCONEV == 2)
#STROKE:
STROKEV <-as.numeric(STROKEV == 2)
#ARTHRITIS:
ARTHLIMIT <-as.numeric(ARTHLIMIT == 2)
#AGE DIAGNOSE WITH DIABETES:
DIABETICAGE <-as.numeric(DIABETICAGE > 01) & (DIABETICAGE < 65)
#ALCOHOL STATUS:
ALCSTAT1 <-as.numeric(ALCSTAT1 == 1) | (ALCSTAT1 == 2) | (ALCSTAT1 == 3)
#NUMBER OF CIGARETTES PER DAY:
CIGSDAY <-as.numeric(CIGSDAY>01) & (CIGSDAY<20)
#SMOKE FREQUENCY:
SMOKFREQNOW <-as.numeric(SMOKFREQNOW == 2) | (SMOKFREQNOW == 3)
#ACTIVITY DURATION:
MOD10DTP <-as.numeric(MOD10DTP == 1) | (MOD10DTP == 2)
#HOURS OF SLEEP PER DAY:
HRSLEEP <-as.numeric(HRSLEEP>01) & (HRSLEEP<08)
#HOW OFTEN DO YOU FEEL SAD:
ASAD <-as.numeric(ASAD>1) & (ASAD<4)
```
#CREATING A NEW DATASET:
```{r}
newdata <-data.frame(REGION,female,educ_nohs,educ_hs,educ_smcoll,educ_as,educ_bach,educ_adv,white,black,asian,native,HEALTH,VISITYRNO,HINOTCOV,ASTHMAEV,CHEARTDIEV,CHOLHIGHEV,HEARTCONEV,STROKEV,ARTHLIMIT,DIABETICAGE,ALCSTAT1,CIGSDAY,SMOKFREQNOW,MOD10DTP,HRSLEEP,ASAD,YEAR,SERIAL,STRATA,PSU,NHISHID,HHWEIGHT,HHX,PX,PERWEIGHT,SAMPWEIGHT,FWEIGHT,ASTATFLG,CSTATFLG,age,BMI)

detach()
attach(newdata)
summary(newdata)
```
#INDEXES:
```{r}
#EDUCATION AND RACE INDEXES:
educ_indx <- as.factor(educ_nohs + 2*educ_hs + 3*educ_smcoll + 4*educ_as + 5*educ_bach + 6*educ_adv)
race_index <-as.factor(white + 2*black)
```
#REGRESSION MODELS:
MODEL1: SEX(female) as the dependent variable
MODEL2: Education as the dependent variable
MODEL3: Race as the dependent variable
```{r}
attach(newdata)

#Linear Model:
Model1<-lm(female ~ ASTHMAEV + CHEARTDIEV + HEARTCONEV + STROKEV + ARTHLIMIT + DIABETICAGE + HEALTH + race_index + REGION + factor(educ_hs) + factor(educ_bach) + factor(educ_adv), data = newdata)
summary(Model1)
coeftest(Model1)
exp(Model1$coefficients)
plot(coef(Model1))

#Logit Model:
Model1a<-glm(female ~ ASTHMAEV + CHEARTDIEV + HEARTCONEV + STROKEV + ARTHLIMIT + DIABETICAGE  + race_index + REGION + factor(educ_hs) + factor(educ_bach) + factor(educ_adv), family = binomial, data = newdata)
summary(Model1a)
coeftest(Model1a)
exp(Model1a$coefficients)
plot(coef(Model1a))

#Probit Model:
Model1b<-glm(female ~ ASTHMAEV + CHEARTDIEV + HEARTCONEV + STROKEV + ARTHLIMIT + DIABETICAGE + HEALTH + race_index + REGION + factor(educ_hs) + factor(educ_bach) + factor(educ_adv), family = binomial(link = "probit"), data = newdata)
summary(Model1b)
coeftest(Model1b)
exp(Model1b$coefficients)
plot(coef(Model1b))
```
```{r}
#Linear Model:
Model2<-lm(female ~ race_index + factor(educ_hs) + factor(educ_bach) + factor(educ_adv) + ASTHMAEV + CHEARTDIEV + HEARTCONEV + STROKEV + ARTHLIMIT + DIABETICAGE + HEALTH)
summary(Model2)
coeftest(Model2)
exp(Model2$coefficients)
plot(coef(Model2))

#Logit Model:
Model2a<-glm(female ~ race_index + factor(educ_hs) + factor(educ_bach) + factor(educ_adv) + ASTHMAEV + CHEARTDIEV + HEARTCONEV + STROKEV + ARTHLIMIT + DIABETICAGE + HEALTH,family = binomial, data = newdata)
summary(Model2a)
coeftest(Model2a)
exp(Model2a$coefficients)
plot(coef(Model2a))

#Probit Model:
Model2b<-glm(female ~ race_index + factor(educ_hs) + factor(educ_bach) + factor(educ_adv) + ASTHMAEV + CHEARTDIEV + HEARTCONEV + STROKEV + ARTHLIMIT + DIABETICAGE + HEALTH, family = binomial(link = "probit"), data = newdata)
summary(Model2b)
coeftest(Model2b)
exp(Model2b$coefficients)
plot(coef(Model2b))
```

```{r}
#Linear Model:
Model3<-lm(female ~ factor(race_index) + factor(educ_hs) + factor(educ_bach) + factor(educ_adv) + ASTHMAEV + CHEARTDIEV + HEARTCONEV + STROKEV + ARTHLIMIT + DIABETICAGE + CIGSDAY + HRSLEEP + VISITYRNO + BMI + factor(REGION))
summary(Model3)
coeftest(Model3)
exp(Model3$coefficients)
plot(coef(Model3))

#Logit Model:
Model3a<-glm(female ~ factor(race_index) + factor(educ_hs) + factor(educ_bach) + factor(educ_adv) + ASTHMAEV + CHEARTDIEV + HEARTCONEV + STROKEV + ARTHLIMIT + DIABETICAGE + CIGSDAY + HRSLEEP + VISITYRNO + BMI, family = binomial, data = newdata)
summary(Model3a)
coeftest(Model3a)
exp(Model3a$coefficients)
plot(coef(Model3a))

#Probit Model:
Model3b<-glm(female ~ factor(race_index) + factor(educ_hs) + factor(educ_bach) + factor(educ_adv) + ASTHMAEV + CHEARTDIEV + HEARTCONEV + STROKEV + ARTHLIMIT + DIABETICAGE + CIGSDAY + HRSLEEP + VISITYRNO + BMI, family = binomial(link = "probit"), data = newdata)
summary(Model3b)
coeftest(Model3b)
exp(Model3b$coefficients)
plot(coef(Model3b))
```

#ANOVA TEST:
```{r}
#MODEL 1 ANOVA:
anova(Model1)
anova(Model1a)
anova(Model1b)

#MODEL 2 ANOVA:
anova(Model2)
anova(Model2a)
anova(Model2b)

#MODEL 3 ANOVA:
anova(Model3)
anova(Model3a)
anova(Model3b)
```